Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-24 paper.tex
spelling

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to imprecise results as the shape of the ligand is not considered.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOFs, that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using the electrostatic energy.
The high-energy configurations are discarded, while the low-energy configurations (and their vicinity) are sampled densely.
The approach~\cite{singhLIG} can also predict the ligand binding sites.

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
The tree can slightly detour from the tunnel and overcome difficult parts (e.g., the tunnel bottleneck) by finding an alternative trajectory nearby.
%It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
%should be allowed, which is controlled by the $\dt$ parameter.
%parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
In case of tunnels not having other tunnels in the close vicinity (which is very common), we recommend to use the double value of the tunnel average width.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was launched with the following parameters:
$\Imax=5000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners, as it resulted in the highest success rates.
%The \RA\ method was also tested without the atom-based metric.
%In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.

@@

As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assume that the ligand can reach the active site.
Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and
the active site was considered as rached if the distance to the geometric center of the ligand was less than 5~\AA.
All paramters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
The proposed planner was able to find trajectories even for this large ligand inside the long tunnels, which is
shown by the traversability rates (Tab.~\ref{tab::4l2l}).
Example of the solution is depicted in Fig.~\ref{fig::inhibitor}.

@@

According to the results of the motion planing, the DCP ligand needs to shrink down to approximately 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
Beyond this point, the ligand can move at larger scales again.
As both the protein and ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}. 
The necessity to scaled-down even more than 0.8, which is a value used by other tools~\cite{cortes2010simulating}, was
shown also in the second experiment, where the trajectory for the inhibitor of the 4L2L protein required scaling down to 0.5 or 0.6.

>>>>

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to imprecise results as the shape of the ligand is not considered.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for trajectory computations of flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

In the case of RRT-based planners, increasing the probability of sampling in difficult regions does not need to bring any advantage
if the growth of the tree is blocked by the obstacles.
%The RRT-based planners cannot cope with the narrow passage problem by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

The shrinking technique has also been used in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOFs, that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using the electrostatic energy.
The high-energy configurations are discarded, while the low-energy configurations (and their vicinity) are sampled densely.
The approach~\cite{singhLIG} can also predict the ligand binding sites.

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
The tree can slightly detour from the tunnel and overcome difficult parts (e.g., the tunnel bottleneck) by finding an alternative trajectory nearby.
%It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
%should be allowed, which is controlled by the $\dt$ parameter.
%parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
In case of tunnels not having other tunnels in the close vicinity (which is very common), we recommend using the double value of the tunnel average width.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was launched with the following parameters:
$\Imax=5000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable a fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners, as it resulted in the highest success rates.
%The \RA\ method was also tested without the atom-based metric.
%In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.

@@

As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assume that the ligand can reach the active site.
Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and
the active site was considered as reached if the distance to the geometric center of the ligand was less than 5~\AA.
All parameters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
The proposed planner was able to find trajectories even for this large ligand inside the long tunnels, which is
shown by the traversability rates (Tab.~\ref{tab::4l2l}).
Example of the solution is depicted in Fig.~\ref{fig::inhibitor}.

@@

According to the results of the motion planning, the DCP ligand needs to shrink down to approximately 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
Beyond this point, the ligand can move at larger scales again.
As both the protein and ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}. 
The necessity to scaled-down even more than 0.8, which is a value used by other tools~\cite{cortes2010simulating}, was
shown also in the second experiment, where the trajectory for the inhibitor of the 4L2L protein required scaling down to 0.5 or 0.6.
